---
title: "Organic Farming Across Europe"
subtitle: "A spatial analysis of organic area as a share of agricultural land"
author: "Christopher Jr Galgo"
date: today
format:
  html:
    toc: true
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    theme: flatly
    fig-width: 8
    fig-height: 7
execute:
  warning: false
  message: false
---

<!--
READ THIS FIRST

This is your second project. It teaches spatial analysis in R.
The new ideas are: sf objects, coordinate reference systems (CRS),
projections, and joining tabular data to geometry.

Two honest warnings from me:

1. I could not run R when I wrote this file, so Eurostat dataset codes
   and column names may have shifted. Section 2 includes a discovery
   step that finds the current code. Treat breakage as normal; every
   analyst debugs API changes.

2. Downloads happen on first render and may take a few minutes.

Render with: quarto render analysis.qmd
-->

## 1. Business question

The EU Farm to Fork strategy targets 25% of agricultural land under
organic production by 2030. Progress varies widely between member
states. This analysis asks: **where does organic farming concentrate,
how fast is each country moving, and which countries will miss the
target on current trends?**

The spatial dimension matters here. Organic adoption clusters
geographically, and clusters suggest shared drivers such as policy
regimes, consumer markets, or agronomic conditions.

## 2. Packages and data discovery

```{r}
#| label: setup

# One-time install (run in the console, not here):
# install.packages(c("eurostat", "sf", "giscoR", "tidyverse",
#                    "scales", "knitr"))

library(eurostat)    # Eurostat data + NUTS boundaries
library(sf)          # simple features: the standard spatial class in R
library(tidyverse)
library(scales)
library(knitr)
```

Eurostat retires and renames datasets. Rather than hard-coding a code
and hoping, find it:

```{r}
#| label: discover
#| eval: false

# Set eval: true above to run this once. It searches dataset titles.
search_eurostat("organic") |>
  select(title, code) |>
  head(20)
```

We use `sdg_02_40`, the SDG indicator for area under organic farming.

```{r}
#| label: load-data

organic_raw <- get_eurostat("sdg_02_40")

# ALWAYS inspect before filtering. You cannot filter columns you have
# not looked at.
glimpse(organic_raw)

# What units and categories does this dataset contain?
organic_raw |>
  count(unit) |>
  kable(caption = "Units available in the dataset")
```

```{r}
#| label: normalise

# Defensive step: eurostat renamed its time column across versions.
# This works either way, and documents the problem for whoever reads
# your code later.
time_col <- intersect(c("TIME_PERIOD", "time"), names(organic_raw))[1]

organic <- organic_raw |>
  rename(year = all_of(time_col)) |>
  mutate(year = as.integer(format(as.Date(year), "%Y"))) |>
  filter(unit == "PC_UAA") |>          # percent of utilised agri. area
  filter(!is.na(values))

# If the unit filter empties the table, the code above is wrong for the
# current data. Check the unit table printed in the previous chunk.
cat("Rows retained:", nrow(organic),
    "| Years:", min(organic$year), "to", max(organic$year), "\n")
```

## 3. Joining data to geometry

```{r}
#| label: geometry

# NUTS level 0 = countries. This call uses the copy bundled inside the
# eurostat package, so it needs no download.
nuts0 <- get_eurostat_geospatial(
  output_class = "sf",
  resolution = "60",
  nuts_level = 0,
  year = 2016
)

# An sf object is a data frame with a geometry column attached.
# Everything you know from dplyr still works on it.
class(nuts0)
```

```{r}
#| label: join

latest_year <- max(organic$year)

map_data <- nuts0 |>
  left_join(
    organic |> filter(year == latest_year),
    by = "geo"                    # NUTS code, e.g. "FR", "DE"
  )

# Report the join quality. Silent join failures are the most common
# error in spatial work, and reviewers look for this check.
n_matched <- sum(!is.na(map_data$values))
cat(n_matched, "of", nrow(map_data), "geographies matched.\n")
```

## 4. Projections

```{r}
#| label: projection

# Raw NUTS data arrive in EPSG:4326 (longitude/latitude on a sphere).
# That projection distorts area badly at European latitudes: Finland
# looks far larger than it is.
st_crs(map_data)$epsg

# EPSG:3035 (ETRS89-LAEA) is the EU's official equal-area projection.
# Use it whenever you map Europe. This single line separates maps that
# look professional from maps that do not.
map_proj <- st_transform(map_data, crs = 3035)
```

## 5. Finding 1: Organic farming concentrates in the Alpine and Nordic band

```{r}
#| label: fig-choropleth
#| fig-cap: "Organic area as a share of utilised agricultural area"

ggplot(map_proj) +
  geom_sf(aes(fill = values), colour = "white", linewidth = 0.2) +
  # Bounding box in projected metres, cropping overseas territories
  coord_sf(xlim = c(2400000, 6400000), ylim = c(1400000, 5500000)) +
  scale_fill_viridis_c(
    option = "viridis", na.value = "grey90",
    labels = label_number(suffix = "%")
  ) +
  labs(
    title = paste("Organic farming share of agricultural area,",
                  latest_year),
    fill = "% of UAA",
    caption = "Source: Eurostat (sdg_02_40). Grey = no data."
  ) +
  theme_void() +
  theme(legend.position = "right")
```

Write your own reading of this map here, in two or three sentences.
Name the leaders, name the laggards, and resist explaining the pattern
before you have tested it. Description first, interpretation second.

**YOUR TURN 1:** Remove the `st_transform()` line and render again.
Compare the two maps. Write one sentence on what changed and why it
matters.

## 6. Finding 2: Distance from the 25% target

```{r}
#| label: fig-gap
#| fig-cap: "Gap to the Farm to Fork 2030 target of 25%"

target <- 25

gap <- organic |>
  filter(year == latest_year, nchar(geo) == 2, geo != "EU27_2020") |>
  mutate(
    gap_to_target = target - values,
    reached = values >= target
  ) |>
  arrange(desc(values))

ggplot(gap, aes(x = values, y = reorder(geo, values), fill = reached)) +
  geom_col() +
  geom_vline(xintercept = target, linetype = "dashed") +
  scale_fill_manual(values = c("FALSE" = "#7c9885", "TRUE" = "#2c5f4a"),
                    guide = "none") +
  labs(
    x = "% of utilised agricultural area", y = NULL,
    title = paste("Most member states sit far below 25% in",
                  latest_year),
    caption = "Dashed line: Farm to Fork 2030 target."
  ) +
  theme_minimal()
```

**YOUR TURN 2:** The country codes are hard to read. Use
`label_eurostat()` from the eurostat package to attach country names,
then redraw this chart.

## 7. Finding 3: Growth rates, not levels, decide who reaches the target

```{r}
#| label: fig-trend
#| fig-cap: "Change in organic share since 2012"

trend <- organic |>
  filter(nchar(geo) == 2, geo != "EU27_2020",
         year %in% c(2012, latest_year)) |>
  select(geo, year, values) |>
  pivot_wider(names_from = year, values_from = values,
              names_prefix = "y") |>
  drop_na() |>
  mutate(
    change = .data[[paste0("y", latest_year)]] - y2012,
    annual_rate = change / (latest_year - 2012)
  )

ggplot(trend, aes(x = y2012,
                  y = .data[[paste0("y", latest_year)]])) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_point(colour = "#2c5f4a", size = 2.5) +
  geom_text(aes(label = geo), nudge_y = 0.9, size = 3) +
  labs(
    x = "Organic share, 2012 (%)",
    y = paste0("Organic share, ", latest_year, " (%)"),
    title = "Countries above the dotted line expanded organic area",
    caption = "Dotted line: no change."
  ) +
  theme_minimal()
```

A simple linear extrapolation shows how few countries reach 25% by 2030
on current trends:

```{r}
#| label: tbl-projection

trend |>
  mutate(
    projected_2030 = .data[[paste0("y", latest_year)]] +
      annual_rate * (2030 - latest_year),
    reaches_target = if_else(projected_2030 >= 25, "Yes", "No")
  ) |>
  arrange(desc(projected_2030)) |>
  select(geo, annual_rate, projected_2030, reaches_target) |>
  mutate(across(where(is.numeric), \(x) round(x, 1))) |>
  head(15) |>
  kable(col.names = c("Country", "Annual change (pp)",
                      "Projected 2030 (%)", "Reaches 25%?"))
```

Linear extrapolation assumes constant growth, which rarely holds.
Adoption curves usually flatten as the willing early adopters convert.
Treat these projections as an optimistic ceiling, not a forecast.

## 8. Recommendations

1. **Target the laggards, not the leaders.** Countries already above
   20% face diminishing returns; the aggregate EU target depends on
   movement in the large, low-share agricultural economies.
2. **Study the fast movers.** Countries with steep growth from a low
   base reveal which policy levers work, and their conditions
   resemble the laggards' more than the leaders' do.
3. **Treat the 25% target as aggregate, not uniform.** Agronomic
   conditions constrain organic conversion; a single national threshold
   for every member state ignores this.

## 9. Limitations

Eurostat reports certified organic area, which excludes low-input
practices that deliver similar environmental outcomes without
certification. Country-level aggregation conceals within-country
variation; NUTS2 analysis would show more. The projections assume
linear growth and no policy change. Association between geography and
adoption does not establish that geography causes adoption; market
access, subsidy design, and farm size all plausibly confound it.

With more time, I would repeat this at NUTS2 level and test whether
neighbouring regions correlate more strongly than distant ones, which
would give the spatial claim statistical support rather than visual
support alone.

<!--
STRETCH GOALS (later, with my help):
1. NUTS2 version using a regional dataset.
2. A spatial autocorrelation test (Moran's I via the spdep package) to
   replace "clusters visibly" with a measurable statistic.
3. Convert to format: dashboard.
-->
